---
title: "Getting started with backend-aware matrices in amatrix"
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
  - albers.css
  - albers.js
  - albers-header.html
vignette: >
  %\VignetteIndexEntry{Getting started with backend-aware matrices in amatrix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::theme_set(ggplot2::theme_minimal())
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

```{r load-amatrix}
library(amatrix)
```

`amatrix` is for matrix-heavy code that already fits the `Matrix` idiom but needs a cleaner path to accelerated execution. You keep ordinary matrix inputs, wrap them once, and then use Matrix-compatible objects that carry backend preferences into linear algebra and model fitting.

This vignette uses one common workload: one design matrix, many response columns. By the end, you will have a backend-aware design matrix, a shared-design QR fit from `many_lm()`, and a compact way to inspect where `amatrix` plans to run each operation.

## What do your inputs look like?

```{r example-data, include = FALSE}
set.seed(42)

n <- 120L
p <- 6L
q <- 8L

X <- scale(
  matrix(rnorm(n * p), nrow = n, ncol = p),
  center = TRUE,
  scale = FALSE
)
colnames(X) <- paste0("feature_", seq_len(p))

beta <- matrix(rnorm(p * q, sd = 0.7), nrow = p, ncol = q)
noise <- matrix(rnorm(n * q, sd = 0.2), nrow = n, ncol = q)
Y <- X %*% beta + noise
colnames(Y) <- paste0("response_", seq_len(q))
```

```{r data-shape}
c(
  observations = nrow(X),
  predictors = ncol(X),
  responses = ncol(Y)
)
```

The design matrix `X` is an ordinary dense R matrix with one row per observation and one column per predictor. The response matrix `Y` has the same number of rows, but each column is a separate regression target that shares the same design.

## What is the quickest end-to-end path?

```{r first-fit}
X_am <- adgeMatrix(X)
fit <- many_lm(X_am, Y, method = "qr", include_fitted = TRUE, include_residuals = TRUE)
dim(coef(fit))
```

`adgeMatrix()` is the main dense-matrix constructor. `many_lm()` then fits all response columns against the same design in one call and returns a coefficient matrix with one column per response.

```{r first-fit-checks, include = FALSE}
qr_ref <- qr.coef(qr(X), Y)
coef_err <- max(abs(as.matrix(coef(fit)) - qr_ref))
r2 <- 1 - fit$rss / colSums(Y^2)

stopifnot(is.finite(coef_err), coef_err < 1e-8)
stopifnot(all(is.finite(fit$rss)), all(is.finite(fit$sigma2)))
stopifnot(all(fit$rss > 0), all(fit$sigma2 > 0))
stopifnot(all(is.finite(r2)), all(r2 > 0.85), all(r2 < 1))
```

## How does `amatrix` decide where to run?

```{r backend-status}
amatrix_backend_status()[, c("name", "available", "precision_modes", "residency_capable")]
```

The status table tells you which backends are registered on the current machine and whether they are usable right now. A backend can be registered but unavailable, in which case the same code still has a predictable CPU fallback.

```{r backend-plan}
plan <- amatrix_backend_plan(X_am, "qr")
plan[c("chosen", "chosen_path", "requested_precision")]
```

This plan is the compact view of a dispatch decision. You can see which backend was chosen, whether the operation will run as a cold or resident path, and which precision contract the object is requesting.

```{r backend-plan-checks, include = FALSE}
status <- amatrix_backend_status()
stopifnot(plan$chosen %in% status$name)
stopifnot(plan$requested_precision %in% c("strict", "fast"))
```

## When does shared-design caching help?

The main reason to use `amatrix` for repeated regression is that the design-matrix factorization can be reused across calls when `X` stays the same. That matters when you fit many batches of responses against one shared design.

```{r cache-demo}
X_cache_am <- adgeMatrix(X)
fit_a <- many_lm(X_cache_am, Y[, 1:4, drop = FALSE], method = "qr", cache = TRUE)
fit_b <- many_lm(X_cache_am, Y[, 5:8, drop = FALSE], method = "qr", cache = TRUE)

c(first_call = fit_a$cache_reused, second_call = fit_b$cache_reused)
```

With a fresh backend-aware copy of `X`, the first call computes and stores the QR factorization for that design. The second call reuses it because the design matrix is unchanged.

```{r cache-demo-checks, include = FALSE}
coef_a_ref <- qr.coef(qr(X), Y[, 1:4, drop = FALSE])
coef_b_ref <- qr.coef(qr(X), Y[, 5:8, drop = FALSE])

stopifnot(isFALSE(fit_a$cache_reused))
stopifnot(isTRUE(fit_b$cache_reused))
stopifnot(max(abs(as.matrix(coef(fit_a)) - coef_a_ref)) < 1e-8)
stopifnot(max(abs(as.matrix(coef(fit_b)) - coef_b_ref)) < 1e-8)
```

## What should you inspect after a fit?

```{r fit-summary}
summary_tbl <- data.frame(
  response = colnames(Y),
  rss = round(fit$rss, 3),
  sigma2 = round(fit$sigma2, 4),
  r2 = round(r2, 3)
)

knitr::kable(summary_tbl, align = "lrrr")
```

`rss` and `sigma2` tell you how much unexplained variation remains in each response. Here the fitted models recover most of the signal because the responses were generated from a shared linear design with modest noise.

```{r fitted-response-plot, echo = FALSE, fig.height = 4, fig.width = 5}
plot(
  as.matrix(fitted(fit))[, 1],
  Y[, 1],
  xlab = "Fitted values",
  ylab = "Observed response_1",
  main = "Shared-design fit for one response",
  pch = 19,
  col = "#1B5E20"
)
abline(0, 1, lty = 2, col = "grey40")
```

The points fall close to the identity line, which matches the high `r2` values from the hidden checks and the small residual variances in the summary table.

## How do you ask for a fast backend?

The CPU path above is the safest default because it runs everywhere. On a machine with an available accelerator backend, the code shape stays the same and you only change the constructor metadata:

```{r fast-backend-example, eval = FALSE}
X_fast <- adgeMatrix(X, mode = "fast")
fit_fast <- many_lm(X_fast, Y, method = "qr", cache = TRUE)

coef(fit_fast)
```

If you are writing library code, the default path is still to wrap once at the boundary and keep the rest of the code generic:

```{r fast-library-path, eval = FALSE}
X_am <- as_adgeMatrix(X, mode = "fast")
fit <- many_lm(X_am, Y, method = "qr", cache = TRUE)
```

That per-object path does not require session-global setters or a hardcoded backend name. `amatrix` will prefer an available fast-capable accelerator automatically and fall back to CPU when none is available.

If a caller wants to flip defaults for one local block instead of one object, use `with_amatrix()`:

```{r with-amatrix-example, eval = FALSE}
with_amatrix(policy = "auto", precision = "fast", {
  X_am <- as_adgeMatrix(X)
  fit <- many_lm(X_am, Y, method = "qr", cache = TRUE)
})
```

Use `amatrix_bind_resident()` only when you need to optimize a particularly hot repeated path.

## Where next?

The next functions to explore are `adgeMatrix()`, `many_lm()`, `amatrix_backend_status()`, and `amatrix_explain()`. If you want to audit a single operation more deeply, start with `amatrix_backend_plan()` and then compare that compact plan to the printed explanation from `amatrix_explain()`.
