---
title: "GPU Acceleration: Zero to First Matmul"
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: lapis
  preset: homage
resource_files:
  - albers.css
  - albers.js
  - albers-header.html
vignette: >
  %\VignetteIndexEntry{GPU Acceleration: Zero to First Matmul}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

```{r load-amatrix}
library(amatrix)
```

Core `amatrix` is CPU-first and always works with no accelerator dependencies. GPU execution comes from optional sister packages that plug in at runtime: `amatrix.mlx` (Apple Silicon via MLX), `amatrix.metal` (macOS), `amatrix.arrayfire` (ArrayFire), and `amatrix.opencl` (OpenCL + CLBlast). This vignette takes you from a fresh install to a matmul running on the GPU, then explains the three things that surprise people first: precision, routing, and residency.

Because the GPU path depends on hardware that is not present on every machine (and never on CRAN's build servers), the GPU-specific chunks below are shown with `eval = FALSE` and their console output written out inline. Copy them into a session with a backend installed and they run as printed.

## How do I install a backend?

The backends ship from the `bbuchsbaum` R-universe. Install core `amatrix` on its own for pure-CPU work, and add exactly one accelerator package for your platform.

```{r install, eval = FALSE}
# Core package (CPU-only, no accelerator dependencies)
install.packages("amatrix", repos = c("https://bbuchsbaum.r-universe.dev", getOption("repos")))

# Apple Silicon GPU via MLX
install.packages("amatrix.mlx", repos = c("https://bbuchsbaum.r-universe.dev", getOption("repos")))
```

Pick the backend package that matches your machine:

| Backend package     | Platform                         | System prerequisite                                                        |
|:--------------------|:---------------------------------|:---------------------------------------------------------------------------|
| `amatrix.mlx`       | macOS arm64 (Apple Silicon)      | Homebrew `mlx-c`, or `MLX_C_PREFIX`; builds a mock bridge without it        |
| `amatrix.metal`     | macOS                            | Xcode Command Line Tools                                                    |
| `amatrix.arrayfire` | any OS with ArrayFire >= 3.8     | ArrayFire runtime (on arm64 macOS it pins to the CPU runtime by default)    |
| `amatrix.opencl`    | unix with OpenCL + CLBlast       | OpenCL driver and CLBlast; GPU probe is env-gated (see below)               |

## Does it just work?

On Apple Silicon, yes. Install `amatrix.mlx` and MLX probing is on by default in every launch mode -- interactive, `Rscript -e`, and `Rscript file.R`. The first time a session needs the GPU, `amatrix` runs a one-shot probe in a disposable child process (so a bad driver crashes the child, never your session) and then routes eligible work to MLX automatically. There is nothing to call.

You will see a one-line note when the package attaches, confirming what is available:

```{r startup-note, eval = FALSE}
library(amatrix)
#> amatrix GPU backends: mlx ready (activates on first use). See amatrix_gpu_status().
```

That line is suppressible the usual way with `suppressPackageStartupMessages(library(amatrix))`, and the automatic probe can be turned off with `options(amatrix.auto_probe = FALSE)` or the environment variable `AMATRIX_MLX_PROBE_GPU=0`.

Everywhere else -- OpenCL, ArrayFire, Metal -- the backends are opt-in, so you turn them on with one call. `amatrix_use_gpu()` walks the preferred order, enables and health-probes the first installed backend, sets the fast-precision default, and prints one confirmation line:

```{r use-gpu, eval = FALSE}
amatrix_use_gpu()
#> amatrix: GPU enabled - opencl backend (float32 'fast' precision, ~1e-4 vs
#> float64; 'strict' float64 stays on CPU). amatrix_gpu_status() for details.
```

`amatrix_use_gpu()` works on Apple Silicon too if you want the explicit confirmation line, or pass `backend = "mlx"` to force a specific one. It returns the enabled backend name invisibly, or `FALSE` if none could be brought up.

## Am I actually on the GPU?

Two functions answer this. `amatrix_gpu_status()` is the "why am I (not) on the GPU?" table -- one row per backend, walking every gate between *installed* and *computing on the device*:

```{r gpu-status, eval = FALSE}
amatrix_gpu_status()
#>     backend       package installed registered available  health
#> 1       mlx   amatrix.mlx      TRUE       TRUE      TRUE healthy
#> 2     metal amatrix.metal     FALSE      FALSE     FALSE unprobed
#> 3 arrayfire amatrix.arr..     FALSE      FALSE     FALSE unprobed
#> 4    opencl amatrix.opencl    FALSE      FALSE     FALSE unprobed
#>                                       reason
#> 1                                       <NA>
#> 2              package amatrix.metal not installed
#> 3          package amatrix.arrayfire not installed
#> 4             package amatrix.opencl not installed
```

Read it left to right: a backend has to be `installed`, then `registered`, then report `available`, then pass a `health` probe before work lands on it. The `reason` column names the first gate that failed -- "package not installed", "opt-in backend; enable with amatrix_use_gpu()", or a driver-specific message from the probe.

For a single operation, `amatrix_explain()` shows the actual dispatch decision: which backend was chosen, why every candidate was accepted or rejected, and what to do about it.

```{r explain-gpu, eval = FALSE}
X <- adgeMatrix(matrix(rnorm(4096 * 4096), 4096), mode = "fast")
amatrix_explain(X, "matmul")
#> â”€â”€ amatrix dispatch: matmul â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€
#>   object:    adgeMatrix [4096 Ă— 4096]  precision=fast  preferred=mlx
#>   residency: host (not GPU-resident)
#>
#> â”€â”€ candidates â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€
#>   â–ş CHOSEN   mlx            reg avail prec cold calib  [cold]
#>     ......   cpu            reg avail prec cold calib
#>
#> â”€â”€ result â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€
#>   chosen: mlx  via cold path (upload + compute)
#> â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€
```

The `mode = "fast"` on the constructor is what makes the object eligible for the GPU at all -- more on that next. `amatrix_explain()` returns the underlying plan invisibly; `amatrix_backend_plan(X, "matmul")` gives you the same decision as a plain list if you want to test against it in code.

## What does "fast" precision actually mean?

GPU backends in `amatrix` compute in **float32 only**. That is the deliberate contract: `"fast"` precision runs on the GPU in single precision, and `"strict"` precision (float64) **always** computes on the CPU reference backend, no matter which GPU is enabled. There is no float64 GPU path to accidentally fall into.

You opt an object into the GPU-eligible path when you build it:

```{r fast-mode, eval = FALSE}
X_fast <- adgeMatrix(A, mode = "fast")          # per object
amatrix_set_default_precision("fast")            # or session-wide
```

The practical consequence is that GPU results match base R to about `1e-4` relative error, not to machine epsilon. This is the cross-backend conformance tolerance the package tests against. So this is expected and correct:

```{r conformance, eval = FALSE}
A <- matrix(rnorm(2048 * 2048), 2048)
gpu <- as.matrix(adgeMatrix(A, mode = "fast") %*% adgeMatrix(A, mode = "fast"))
cpu <- A %*% A

max(abs(gpu - cpu)) / max(abs(cpu))
#> [1] 7.2e-05

isTRUE(all.equal(gpu, cpu))
#> [1] "Mean relative difference: 3.1e-05"   # float32, as designed

isTRUE(all.equal(gpu, cpu, tolerance = 1e-4))
#> [1] TRUE
```

If you need bit-for-bit agreement with base R, use `mode = "exact"` (or leave precision at the `"strict"` default) and the work stays on the CPU. Reach for `"fast"` when a `1e-4` tolerance is acceptable for the numerics you are doing -- which, for most large matmuls, factorizations, and iterative solvers, it is.

## Why did my small matrix stay on the CPU?

Because that is faster, and `amatrix` knows it. Moving data to the GPU costs an upload; you only win that cost back when the compute is large enough to amortize it. Each backend declares calibrated size thresholds (for example, MLX declines `crossprod` below a 2048 leading dimension) and simply refuses the op below them, so small matrices route to the tuned CPU BLAS.

So seeing `cpu` chosen for a small input is the system working, not failing:

```{r small-cpu, eval = FALSE}
small <- adgeMatrix(matrix(rnorm(64 * 64), 64), mode = "fast")
amatrix_explain(small, "crossprod")
#> ...
#> â”€â”€ result â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€
#>   chosen: cpu  via cold path (upload + compute)
#>
#> â”€â”€ suggestions â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€
#> * 'mlx' skipped by calibration: 4,096 elements < threshold 4,194,304.
#>   â†’ Use a larger matrix, or re-run amatrix_calibrate() if the
#>     threshold seems wrong for your hardware.
#> â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€â”€
```

`amatrix_explain()` spells out the reason ("skipped by calibration") so you never have to guess whether a CPU result means the GPU is broken. It isn't -- the input was just too small to be worth it. The `performance` vignette (`vignette("performance")`) covers where the crossover lies and how to recalibrate for your hardware.

## What happens when the GPU fails?

It falls back to the CPU and keeps going. Every runtime GPU failure degrades to the CPU reference path, marks the backend's health, and records an entry you can inspect:

```{r fallback-log, eval = FALSE}
amatrix_fallback_log()
#>             timestamp     op from_backend to_backend                     reason
#> 1 2026-07-01 10:14:22 matmul          mlx        cpu   kernel launch failed: ...
```

A non-empty fallback log means a backend claimed an op it could not actually execute -- worth reporting, but your results are still correct because the CPU produced them. Clear it with `amatrix_fallback_log_reset()` when you want to isolate a fresh run.

A more severe failure -- a crash during the first-use probe -- is contained in the child process and disables that backend for the rest of the session with an actionable reason. You will see it in `amatrix_gpu_status()`:

```{r degraded-status, eval = FALSE}
amatrix_gpu_status()[, c("backend", "available", "health", "reason")]
#>   backend available   health                              reason
#> 1     mlx     FALSE unhealthy  isolated GPU probe crashed; disabled for session
```

The troubleshooting loop is always the same: read the `reason` column. It routes you to the fix -- install the package, call `amatrix_use_gpu()`, set the probe env var, use a bigger matrix, or file the driver crash.

## The one sharp edge: resident handles alias the GPU buffer

Ordinary `amatrix` objects are copy-on-write safe, exactly like base R matrices. Assigning and then modifying a copy never touches the original:

```{r cow-safe, eval = FALSE}
Y <- X
Y[1, 1] <- 999
X[1, 1]        # unchanged
#> [1] 0.418
```

`resident_handle()` is the deliberate exception. It is a mutable wrapper used by iterative algorithms to avoid per-step object allocation, and its in-place operators (`am_sweep_inplace()`, `am_ewise_inplace()`) mutate the device buffer directly. When you build a handle from a source that is **already GPU-resident**, the handle *reuses that same buffer* rather than re-uploading -- so an in-place op through the handle also mutates the source object and any aliases pointing at that buffer:

```{r resident-alias, eval = FALSE}
X <- adgeMatrix(A, mode = "fast")
X <- amatrix_bind_resident(X, "mlx")   # X now owns a resident buffer

h <- resident_handle(X)                # reuses X's buffer -- shared, not copied
am_ewise_inplace(h, 2, "*")            # doubles the buffer in place

# X sees the change too -- reference semantics, by design
as.matrix(X)[1, 1] == 2 * A[1, 1]
#> [1] TRUE
```

This is intentional: sharing the buffer is the whole point of a resident handle on a hot path. If you want an independent copy to mutate, build the handle from a **non-resident** source (a plain matrix, or an `adgeMatrix` that has not been bound resident), in which case `resident_handle()` uploads a fresh buffer that it owns exclusively. When in doubt, keep in-place handle work off of objects you still intend to read.

## Where next?

- `vignette("amatrix")` -- the getting-started workflow (`adgeMatrix()`, `many_lm()`)
- `vignette("performance")` -- when the GPU actually wins, and how to calibrate
- `?amatrix_use_gpu` and `?amatrix_gpu_status` -- enablement and diagnostics
- `?amatrix_explain` -- per-operation dispatch reasoning
- `?resident_handle` -- the mutable device-resident path for iterative code
