---
title: "amatrix Benchmark Report"
subtitle: "Benchmark bundle generated by `tools/benchmark-regression.R`"
execute:
  echo: false
  warning: false
  message: false
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    css: benchmark-report.css
    theme: cosmo
    page-layout: full
  pdf:
    toc: true
    toc-depth: 2
    documentclass: scrartcl
    geometry:
      - margin=0.85in
    colorlinks: true
    include-in-header: benchmark-report-preamble.tex
    fig-pos: "H"
---

```{r}
#| label: setup
#| include: false

report <- readRDS("report-data.rds")

metric_value <- function(name) {
  idx <- match(name, report$tables$metrics$metric)
  if (is.na(idx)) return(NA_character_)
  as.character(report$tables$metrics$value[[idx]])
}

render_table <- function(df, digits = 3, max_rows = 12L) {
  if (nrow(df) == 0L) {
    cat("_None._\n")
    return(invisible(NULL))
  }

  abbreviate_table <- function(df, kind) {
    if (!knitr::is_latex_output()) {
      return(df)
    }

    truncate_text <- function(x, width = 44L) {
      x <- as.character(x)
      too_long <- !is.na(x) & nchar(x) > width
      x[too_long] <- paste0(substr(x[too_long], 1L, width - 3L), "...")
      x
    }

    out <- switch(
      kind,
      policy_notes = df[c("area", "policy", "detail")],
      backend_overview = df[c(
        "requested_backend", "total_cells", "executed_cells",
        "accelerated_cells", "cpu_fallback_cells", "incident_cells",
        "regression_cells"
      )],
      suite_overview = df[c(
        "suite", "ops", "requested_backends", "total_cells",
        "executed_cells", "non_cpu_executed"
      )],
      op_coverage = df[c(
        "suite", "op", "requested_backends", "executed_backends",
        "variants", "cells", "executed_cells"
      )],
      backend_status = df[c(
        "name", "available", "health", "precision_modes",
        "residency_capable", "capabilities"
      )],
      regressions = df[c(
        "op", "size_label", "variant", "requested_backend",
        "dispatch_backend", "median_ms", "baseline_ms",
        "ratio_vs_baseline", "regression_pct"
      )],
      warm_pairs = df[c(
        "op", "size_label", "requested_backend", "cold_ms",
        "warm_variant", "warm_ms", "cold_to_warm_gain",
        "warm_vs_cold_ratio", "ms_saved"
      )],
      acceleration = df[c(
        "op", "size_label", "variant", "requested_backend",
        "dispatch_backend", "median_ms", "cpu_reference_ms",
        "speedup_vs_cpu", "ratio_vs_baseline"
      )],
      incidents = df[c(
        "op", "size_label", "variant", "requested_backend",
        "status", "error_message"
      )],
      fallbacks = df[c(
        "op", "size_label", "variant", "requested_backend",
        "dispatch_backend", "median_ms", "cpu_reference_ms",
        "requested_support_reason"
      )],
      df
    )

    rename_map <- list(
      policy_notes = c(
        detail = "detail"
      ),
      backend_overview = c(
        requested_backend = "backend",
        total_cells = "total",
        executed_cells = "exec",
        accelerated_cells = "accel",
        cpu_fallback_cells = "fallback",
        incident_cells = "incidents",
        regression_cells = "regress"
      ),
      suite_overview = c(
        requested_backends = "req_backends",
        total_cells = "total",
        executed_cells = "exec",
        non_cpu_executed = "gpu_exec"
      ),
      op_coverage = c(
        requested_backends = "req_backends",
        executed_backends = "exec_backends",
        executed_cells = "exec"
      ),
      backend_status = c(
        available = "avail",
        precision_modes = "precision",
        residency_capable = "resident",
        capabilities = "capabilities"
      ),
      regressions = c(
        size_label = "size",
        requested_backend = "req",
        dispatch_backend = "disp",
        median_ms = "ms",
        baseline_ms = "base_ms",
        ratio_vs_baseline = "ratio",
        regression_pct = "reg_pct"
      ),
      warm_pairs = c(
        size_label = "size",
        requested_backend = "backend",
        warm_variant = "warm",
        cold_to_warm_gain = "gain",
        warm_vs_cold_ratio = "ratio",
        ms_saved = "saved_ms"
      ),
      acceleration = c(
        size_label = "size",
        requested_backend = "req",
        dispatch_backend = "disp",
        median_ms = "run_ms",
        cpu_reference_ms = "cpu_ms",
        speedup_vs_cpu = "speedup",
        ratio_vs_baseline = "base_ratio"
      ),
      incidents = c(
        size_label = "size",
        requested_backend = "backend",
        error_message = "error"
      ),
      fallbacks = c(
        size_label = "size",
        requested_backend = "req",
        dispatch_backend = "disp",
        median_ms = "run_ms",
        cpu_reference_ms = "cpu_ms",
        requested_support_reason = "reason"
      )
    )

    if (!is.null(rename_map[[kind]])) {
      for (old_name in names(rename_map[[kind]])) {
        names(out)[names(out) == old_name] <- rename_map[[kind]][[old_name]]
      }
    }

    for (col_name in intersect(c("req_backends", "exec_backends", "capabilities", "error", "reason", "detail", "policy"), names(out))) {
      out[[col_name]] <- truncate_text(out[[col_name]])
    }

    out
  }

  df <- abbreviate_table(df, attr(df, "report_kind"))
  df <- utils::head(df, max_rows)
  numeric_cols <- vapply(df, is.numeric, logical(1))
  for (name in names(df)[numeric_cols]) {
    df[[name]] <- round(df[[name]], digits = digits)
  }

  fmt <- if (knitr::is_latex_output()) "latex" else "html"
  table_out <- knitr::kable(
    df,
    format = fmt,
    booktabs = knitr::is_latex_output(),
    longtable = FALSE,
    linesep = "",
    escape = TRUE,
    table.attr = if (!knitr::is_latex_output()) 'class="report-table"' else NULL
  )
  if (knitr::is_latex_output()) {
    table_out <- paste(
      "\\begingroup\\scriptsize",
      "\\setlength{\\tabcolsep}{3.5pt}",
      table_out,
      "\\endgroup",
      sep = "\n"
    )
  }
  cat(table_out, sep = "\n")
  invisible(NULL)
}

report_table <- function(df, kind, digits = 3, max_rows = 12L) {
  attr(df, "report_kind") <- kind
  render_table(df, digits = digits, max_rows = max_rows)
}

plot_block <- function(name, caption) {
  path <- file.path("plots", report$plots[[name]])
  if (!file.exists(path)) {
    cat("_Plot unavailable._\n")
    return(invisible(NULL))
  }
  knitr::include_graphics(path)
  if (knitr::is_latex_output()) {
    cat(sprintf("\n\\emph{%s}\n", caption))
  } else {
    cat(sprintf("\n<div class=\"fig-caption\">%s</div>\n", caption))
  }
}

render_code_block <- function(code, language = "bash") {
  cat(sprintf("```%s\n%s\n```\n", language, code))
}
```

::: {.content-visible when-format="html"}
::: {.hero}
::: {.hero-kicker}
`r format(report$metadata$created_at, "%Y-%m-%d %H:%M:%S %Z")`
:::

# Benchmark bundle at a glance

The harness produced `r metric_value("executed_cells")` executed cells with `r if (isTRUE(report$baseline_present)) report$snippets$baseline_path else "no active baseline"` as the reference path. This report turns the bundle into one readable audit surface: regressions, warm-state reuse, routing behavior, reproducibility commands, and the exact files emitted to disk.
:::

::: {.metrics-grid}
::: {.metric-card}
::: {.metric-label}
Executed cells
:::
::: {.metric-value}
`r metric_value("executed_cells")`
:::
:::

::: {.metric-card}
::: {.metric-label}
Regressions >20%
:::
::: {.metric-value}
`r metric_value("regressions_gt_20pct")`
:::
:::

::: {.metric-card}
::: {.metric-label}
Incidents
:::
::: {.metric-value}
`r metric_value("incidents")`
:::
:::

::: {.metric-card}
::: {.metric-label}
CPU fallbacks
:::
::: {.metric-value}
`r metric_value("cpu_fallbacks")`
:::
:::
:::
:::

::: {.content-visible when-format="pdf"}
## Snapshot

- Generated: `r format(report$metadata$created_at, "%Y-%m-%d %H:%M:%S %Z")`
- Host: `r report$metadata$hostname`
- Platform: `r report$metadata$platform`
- R version: `r report$metadata$r_version`
- Baseline: `r if (isTRUE(report$baseline_present)) report$baseline_path else "not found"`
- Executed cells: `r metric_value("executed_cells")`
- Regressions >20%: `r metric_value("regressions_gt_20pct")`
- Incidents: `r metric_value("incidents")`
- CPU fallbacks: `r metric_value("cpu_fallbacks")`
:::

## Run snapshot

This report only covers the suites, operations, and backends that were actually requested for this bundle. It is not a package-wide claim. If an algorithm or backend does not appear in the coverage tables below, it was not benchmarked in this run.

```{r}
#| results: asis
report_table(report$tables$suite_overview, "suite_overview", digits = 0)
```

### Op coverage

```{r}
#| results: asis
report_table(report$tables$op_coverage, "op_coverage", digits = 0, max_rows = 24)
```

## Policy notes

```{r}
#| results: asis
report_table(report$tables$policy_notes, "policy_notes", digits = 0, max_rows = 12)
```

Standalone exploratory scripts such as `tools/benchmark-kmeans.R`, `tools/benchmark-cancor.R`, and `tools/benchmark-krr.R` are not folded into this bundle unless their operations are registered in `tools/benchmark-regression.R`.

- Host: `r report$metadata$hostname`
- Platform: `r report$metadata$platform`
- R version: `r report$metadata$r_version`
- Baseline: `r if (isTRUE(report$baseline_present)) report$baseline_path else "not found"`

```{r}
#| results: asis
report_table(report$tables$backend_overview, "backend_overview", digits = 0)
```

### Backend status snapshot

```{r}
#| results: asis
report_table(report$tables$backend_status, "backend_status", digits = 0, max_rows = 12)
```

## Reproduce this bundle

The report should show exactly how to reproduce the benchmark bundle and how to re-render it into both human-facing formats.

`r if (!is.null(report$snippets$arrayfire_debug)) "Because this bundle includes ArrayFire on Apple Silicon, the direct Rscript launcher is the supported reproduction path here." else ""`

### Benchmark run command

```{r}
#| results: asis
render_code_block(report$snippets$run_bundle, "bash")
```

### Render HTML

```{r}
#| results: asis
render_code_block(report$snippets$render_html, "bash")
```

### Render PDF

```{r}
#| results: asis
render_code_block(report$snippets$render_pdf, "bash")
```

```{r}
#| results: asis
if (!is.null(report$snippets$arrayfire_debug)) {
  cat("### ArrayFire diagnostics\n\n")
  render_code_block(report$snippets$arrayfire_debug, "bash")
}
```

## Read the bundle in R

```{r}
#| results: asis
render_code_block(report$snippets$read_bundle_r, "r")
```

### Summary schema

```{r}
#| results: asis
render_code_block(report$snippets$summary_schema, "text")
```

## Baseline drift

This section answers the question, "what changed versus the saved benchmark baseline?" The plot shows the biggest absolute drifts, while the table below lists only rows that crossed the 20% regression threshold.

```{r}
#| results: asis
plot_block("baseline", "Largest runtime drifts relative to baseline.csv. Values above 1.0 are slower than baseline; values below 1.0 are faster.")
```

```{r}
#| results: asis
report_table(report$tables$regressions, "regressions", digits = 3)
```

## Warm-state reuse

The benchmark harness measures both cold and warmed execution. The plot ranks the best cold-to-warm improvements, which is the quickest way to see where caching or device residency is paying off.

```{r}
#| results: asis
plot_block("warm", "Cold-to-warm gains for the strongest reusable paths in this run.")
```

```{r}
#| results: asis
report_table(report$tables$warm_pairs, "warm_pairs", digits = 3)
```

## Routing and acceleration

The routing plot shows whether requested backends actually executed, fell back to CPU, or never ran. If any non-CPU rows executed on their requested backend, the speedup plot summarizes those wins here as well.

```{r}
#| results: asis
plot_block("routing", "Routing outcomes grouped by requested backend.")
```

```{r}
#| results: asis
plot_block("speedup", "Observed speedup versus the CPU baseline for rows that truly accelerated.")
```

```{r}
#| results: asis
report_table(report$tables$acceleration, "acceleration", digits = 3)
```

## Incidents and fallbacks

```{r}
#| results: asis
report_table(report$tables$incidents, "incidents", digits = 3)
```

```{r}
#| results: asis
report_table(report$tables$fallbacks, "fallbacks", digits = 3)
```

## Artifact manifest

The report bundle is meant to be readable without opening the raw CSVs, but every generated file is still present in the directory beside this report.

- `raw-results.csv`: raw per-cell timing data
- `summary.csv`: decorated summary with routing and performance states
- `regressions.csv`: rows slower than baseline by more than 20%
- `warm-ratios.csv`: matched cold/warm or cold/resident comparisons
- `routing-summary.csv`: backend/routing counts
- `benchmark-summary.md`: plain markdown summary
- `benchmark-report.qmd`: Quarto source for HTML and PDF
- `plots/`: chart assets used in the report
